Utilizing Parallel Workers: LLVM's Vectorization Plan

### 2024-12-04 paper/paper.tex
paper

\begin{document}
\title{Utilizing Parallel Workers: \\LLVM's Vectorization Plan}
\author{Jonas Fritsch}
\affiliation{%
  \institution{Technical University of Munich}
}
\email{jonas.fritsch@tum.de}

\begin{abstract}
  Modern SIMD processors provide various vector registers and ISAs for programmers to utilize. 
  However, manually vectorizing code can be time-consuming, so compilers look to auto-vectorize 
  code for better performance. LLVM, as a widely used compiler toolchain, could only auto-vectorize 
  basic scalar code. Intel proposed a comprehensive refactoring of the underlying system to enable 
  LLVM's auto-vectorization to support more complex optimizations. The suggested Vectorization Plan 
  (VPlan) architecture would be more modular and scalable. Since then, significant efforts have been 
  made to implement this new system. Today, VPlan is already actively involved in vectorizing 
  innermost loops. Other areas, such as outer loop vectorization, are also in development.
\end{abstract}

@@

With these limitations in mind, Intel started an ongoing refactorization effort to migrate LLVM's
auto-vectorization pipeline to utilize a more abstract Vectorization Plan 
(VPlan)~\cite{llvmextloopvec,llvmvplan}. The final goal would be to unite LLVM's auto-vectorization
in a single flexible system capable of optimizing SLPs, inner, and nested loops with complex 
control flows. The auto-vectorization pass would create and transform multiple different 
VPlans, each modeling a different vectorization approach, abstracted away from the underlying LLVM IR.
Those VPlans are compared against each other based on a cost model, and finally, the best one, which
might also not vectorize at all, is materialized into the LLVM IR.

@@

@@

\section{Background}
\label{sec:background}
In general, vectorization techniques can be separated into different categories:

\subsection{Inner Loop Vectorization}
Innermost loops are loops whose body's control flow does not contain any more loops, bound or unbound.
Vectorization of these loops is more straightforward as the only option to vectorize is over 
the loop's induction variable.

@@

\subsection{Super-Word Parallelism (SLP) Vectorization}
SLP Vectorization vectorizes similar independent instructions. A compiler might have, for example, 
fully unrolled a smaller loop in a previous code simplification pass. The resulting similar scalar 
instructions might then be vectorized later by an SLP vectorizer.

\subsection{Outer Loop Vectorizaton}
Outer loop vectorization focuses on control flow with nested loops. Such loops often present
more challenges for auto-vectorization, as one could vectorize over the outer loop,
inner loop, or a mix of both, representing a hybrid/multi-dimensional vectorization approach. 
Carrying out and comparing costs of such different vectorization approaches can be complex 
and require a flexible system like, for example, VPlans.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.39\textwidth]{images/outer-loop-vec.png}
  \caption{A code snippet of a nested loop where it would be beneficial to vectorize over the 
  outer loop. The inner loop has a trip count that is too low (6) to vectorize. Additionally,
  the memory access pattern of \texttt{y} would be consecutive compared to scattered when
  vectorizing over \texttt{i} instead of \texttt{j}.}
  \label{fig:outer-loop-vec}
\end{figure}

\subsection{Function Vectorization}
Function vectorization refers to the vectorization of entire functions. A loop might contain a
non-inlined function call, passing data related to the iteration variable. During auto-vectorization, 
this call could be replaced with a call to a newly constructed function with the same semantic 
control flow but operating on multiple values at once through vector registers instead.

\subsection{Vectorization Similarities}
All of the above-mentioned vectorization techniques can be brought into relation with each other. 
For example, function vectorization can be transformed into loop vectorization by creating a loop 
around the function body and vectorizing it and vice versa. The same applies for SLP Vectorization 
and loop vectorization when unrolling a loop.

\subsection{Vectorization Legality}
Not all loops can be vectorized. There are various factors that can ultimately hinder any
vectorization attempts.

@@

Not all data dependencies directly hinder any vectorization. One exception can 
sometimes be made for reduction idioms, as seen in the inner loop in 
Figure~\ref{fig:outer-loop-vec}, where \texttt{sum} would be a reduction variable.

Additionally, the control flow should be primarily straight-line code. While most if-else 
constructs can be flattened to always execute both branches along with masking, the same is not 
valid for switch-case statements or too complex control flows.

@@

Additionally, the vectorized code size will always be larger than the scalar equivalent. This is
due to many factors: (1) The transformed control flow often includes more instructions.
(2) As the loop trip count might not be a perfect multiple of the vectorization factor, a scalar
epilogue/remainder of the loop might be used to handle the tail iterations. (3) Various runtime
checks might be inserted throughout the vectorized control flow to handle, for example, pointers 
to overlapping memory regions, in which case the scalar version of the loop is executed.

@@

\subsection{History}
LLVM implements two different vectorizers for auto-vectorization, LoopVectorize (LV)
and SLPVectorize. These vectorizers are a part of LLVM's middle-end and run after module 
simplification has taken place. As VPlan still needs to be integrated into SLP-Vectorization, we 
will only focus on explaining the original LoopVectorize structure.

LoopVectorize was designed to focus only on vectorization of innermost loops~\cite{llvmintrvplan}. 
When starting with an initial scalable loop in IR, it would first run a legality phase to check 
whether  and how it is legal to vectorize the loop. This phase would produce artifacts/maps, such as if 
and what runtime aliasing checks would be required. Then, the cost model ran next, creating more 
artifacts based on the scalar IR and legality artifacts. These cost model artifacts would 
associate different vectorization factors with a cost. Lastly, the transformation phase would vectorize 
the underlying IR in one go based on all the produced artifacts.

As a potential solution, Intel proposed a long-term effort to refactor LLVM's whole auto-vectorization
using a system called Vectorization Plan (VPlan)~\cite{llvmextloopvec}.

The goal was to unite any auto-vectorization systems (SLP, Inner-/Outer-Loop) into a single flexible
system~\cite{llvmintrvplan,llvmvplanstate}, as seen in Figure~\ref{fig:vplan-future}. This system would 
use VPlans to represent different 
vectorization approaches abstracted away from the underlying IR. Each VPlan would 
calculate its cost and transform itself to modify the final IR. At first, the Legality phase (yellow) 
would check if the scalar code would be legal to vectorize at all, and if so, initial VPlans would 
be constructed. In the following Planning phase (magenta), these initial VPlans would be optimized 
using various VPlan-to-VPlan transformations~\cite{llvmouterloopstatus}. 
The VPlan's cost model would supervise those transformations. Finally, the best VPlan would be chosen 
and materialized/executed, modifying the underlying IR.

\begin{figure}
  \centering
  \includegraphics[width=0.475\textwidth]{images/vplan-future.png}
  \caption{The planned future architecture for VPlan. The VPlan optimization would
  work as iteratively applying VPlan-to-VPlan transformations.}
  \label{fig:vplan-future}
\end{figure}

\subsection{State of VPlan (LLVM 19.1.4)}
\emergencystretch 6em

In its current state, VPlan is deeply integrated into LV.
When auto-vectorizing, LV starts by checking all the different loops in the IR.
If it finds one, it will try to vectorize it (\texttt{LoopVectorizePass::\allowbreak processLoop}).
Several factors might hinder vectorization for a given loop: 
(1) If the loop was already vectorized before, it will not be vectorized again. 
(2) The Legality phase checks if and how it is legal to vectorize the scalar
loop. (3) The loop's trip count might be too small to make vectorization profitable. (4) The cost 
of runtime checks, such as pointer aliasing, can be too high.

The initial VPlans are constructed based on the legality checks and the legacy LV cost model's 
cost decisions referencing the scalar IR (\texttt{LoopVectorizationPlanner::\allowbreak plan}). 
After initial VPlan construction, various VPlan-to-VPlan 
transformations are applied to optimize it 
(\texttt{LoopVectorizationPlanner::\allowbreak buildVPlansWithVPRecipes}).

Then, the optimal vectorization factor (VF) is calculated based on the set of VPlans and their 
representative cost (\texttt{LoopVectorizationPlanner::\allowbreak cost}), the calculation of 
which still involves the legacy LV cost model.
The best VPlan is picked based on the optimal VF. A VF of 1 means it is not profitable to vectorize 
the loop; however, interleaving/unrolling can still modify the underlying IR. Afterwards, the 
Interleave Count (IC) is determined 
(\texttt{LoopVectorizationCostModel::\allowbreak selectInterleaveCount}).

The best VPlan is then executed (\texttt{LoopVectorizationPlanner::\allowbreak executePlan}). In the 
case of vectorization (VF $>1$), LV will also directly try to vectorize the epilogue loop.

\subsubsection{VPlan Structure}
To better understand the structure of a VPlan, we will use the
simple conditional loop from Figure~\ref{fig:inner-loop-vec} as an example.

Figure~\ref{fig:inner-loop-scalar-ir} shows the scalar IR of the main loop basic blocks \texttt{for.body},
\texttt{if.then}, and \texttt{for.inc}. Figure~\ref{fig:inner-loop-vplan-hcfg-body} shows the equivalent
hierarchical control-flow-graph (CFG) of a VPlan modeling a vectorization approach for this 
scalar loop. Other basic blocks like the scalar epilogue loop and its predication check (is the final
vector trip count equal to the original trip count) are omitted. See Figure~\ref{fig:inner-loop-vplan-hcfg}
in Appendix~\ref{appendix-figures} for the fully modeled HCFG.

%TODO: Combine inner-loop-scalar-ir figure and inner-loop-vplan-hcfg into a single
%       double column figure.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.475\textwidth]{images/inner-loop-vplan-hcfg-loop-body-color.png}
  \caption{The vectorized loop section of the hierarchical CFG of a VPlan modeling the conditional loop
  depicted in Figure~\ref{fig:inner-loop-vec}. Some predefined values are
  \texttt{vp<\%0>} = VF * UF; \texttt{vp<\%1>} = vector-trip-count; \texttt{vp<\%2>} = original trip-count.}
  \label{fig:inner-loop-vplan-hcfg-body}
\end{figure}

The different colored parts in the scalar IR map to the modeled VPlan region with the same color.
The only new part that VPlan models in this example is handling the scalar epilogue loop (purple).

The hierarchical CFG of a VPlan helps abstract it away from the underlying IR. The nodes of this CFG
are either \texttt{VPBasicBlocks}, containing a sequence of one or more \texttt{VPRecipes}, or
\texttt{VPRegionBlocks}, modeling a single-entry-single-exit subgraph.
In the example, \texttt{vector loop} is a region block. All other blocks are basic blocks.

\texttt{VPRegionBlocks} can indicate that their contents are to be replicated a certain number of times
to model more complex control flow for different VF-UF combinations more efficiently in a single VPlan. 
In Figure~\ref{fig:inner-loop-vplan-hcfg-body}, \texttt{vector loop} is not to be replicated (\texttt{<x1>}).

Within our two \texttt{VPBasicBlocks} are lists of recipes (lines of instructions). A recipe
models one or more instructions in the executed IR and can be based on one or more instructions
from the input IR (ingredients). Note that a recipe might also model an instruction with no reference to any input
IR (in our example, this would be the inversion of our vector mask). There are different types of 
recipes. \texttt{VPReplicateRecipe} will either \texttt{CLONE} exactly one instruction from the input IR
into exactly one instruction in the output IR, or \texttt{REPLICATE} an instruction a certain number
of times into output IR (typically found in replicator \texttt{VPRegionBlocks}). Different 
\texttt{VPWidenRecipes} exist to transform/\texttt{WIDEN} any input IR instruction into a vectorized 
version of it. The last remaining recipe type in our example is \texttt{VPInstruction}. This recipe
\texttt{EMIT}s new instructions into the output IR that have no representative in the input IR (no
ingredients).

Furthermore, VPlan recipes can also model definitions and usages of different values. In our example,
the comparison mask \texttt{vp<\%6>} is a \texttt{VPValue} used by three \texttt{VPUsers}.

@@

\subsubsection{VPlan and Integration with LV}
Setting up and integrating the VPlan structure into LV has seen tremendous effort over the last 
few years. Florian Hahn presented future directions for 
VPlan~\cite{llvmvplanupdate}, which mostly revolve around transforming more LV-based decisions into
explicit VPlan-to-VPlan transformations. Additionally, the legacy LV cost model still influences 
many cost model-related decisions. However, the transition to a VPlan-based cost model has already 
started.

@@

One such project that also tries to enable auto-vectorization support for inner-, outer-loop, and 
whole-function vectorization is the region vectorizer from the University Saarland~\cite{rv}.
This region-vectorizer implements partial control-flow linearization~\cite{10.1145/3192366.3192413}
and a divergence analysis that can handle unstructured CFGs. There has been an effort to partially
implement this control-flow linearization and divergence analysis into LLVM's 
upstream~\cite{rvproposal,rvproposaldep}.

@@

\section{Summary and Future Work}
\label{sec:summary}
In conclusion, VPlan has been in active development for several years now. When compiling with LLVM, 
VPlans model initial vectorization approaches, keep track of different optimization decisions, 
and finally, the best VPlan executes to transform the underlying IR. 
However, some VPlan-to-VPlan transformations and a VPlan-based cost model must be included before the 
legacy LV code can be discarded entirely.

Outer loop vectorization support via VPlan is still experimental, separated from the inner loop
LV integration, and runs on a divergent experimental path (\texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path}).
In the near future, development on outer loop vectorization will continue, and merging 
it with LV's VPlan path is planned.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\appendix

\section{Appendix: Additional Figures}\label{appendix-figures}

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{images/inner-loop-vplan-hcfg.png}
  \caption{Full hierarchical CFG of a VPlan modelling the vectorized version of the loop 
  depicted in Figure~\ref{fig:inner-loop-vec} as of LLVM 19.1.4.}
  \label{fig:inner-loop-vplan-hcfg}
\end{figure*}

>>>>

\begin{document}
\title[State of LLVM's Vectorization Plan]{Utilizing Parallel Workers: LLVM's Vectorization Plan}
\author{Jonas Fritsch}
\affiliation{%
  \institution{Technical University of Munich}
}
\email{jonas.fritsch@tum.de}

\begin{abstract}
  Modern SIMD processors provide various vector registers and ISAs for programmers to utilize. 
  However, manually vectorizing code can be time-consuming, so compilers look to auto-vectorize 
  code for better performance. LLVM, as a widely used compiler toolchain, could only auto-vectorize 
  basic scalar code. Intel proposed a comprehensive refactoring of the underlying system to enable 
  LLVM's auto-vectorization to support more complex optimizations. The suggested Vectorization Plan 
  (VPlan) architecture would be more modular and scalable. Since then, significant efforts have been 
  made to implement this new system. Today, VPlan is already actively involved in vectorizing 
  innermost loops. Other areas, such as outer loop vectorization, are also in development.
  In this paper we give an introduction into VPlan and summarize its latest development results.
\end{abstract}

@@

With these limitations in mind, Intel started an ongoing refactorization effort to migrate LLVM's
auto-vectorization pipeline to utilize a more abstract Vectorization Plan 
(VPlan)~\cite{llvmextloopvec,llvmvplan}. The final goal would be to unite LLVM's auto-vectorization
in a single flexible system capable of optimizing SLPs, inner, and nested loops with complex 
control flow.

The auto-vectorization pass would create multiple different VPlans, based on initial legality 
checks. Each VPlan represents a different vectorization approaches, by modeling the
underlying scalar control flow with its own abstracted vectorized version. Different VPlan-to-VPlan
transformations can then optimize these abstracted control flows based on a cost model.
After that, the best VPlan, and with it the best vectorization approach, is chosen by the cost model.
This VPlan is then executed, by going over its own vectorized control flow representation and
materializing it back into the underlying LLVM IR.

@@

The rest of this paper is organized as follows. We first explain different vectorization
strategies and vectorization constraints in general and describe the original auto-vectorization
approach in LLVM. Section~\ref{sec:vplan} then introduces the new VPlan approach, focusing on
its most important individual components and gives an outlook on future developments. In 
Section~\ref{sec:relatedwork} we highlight related work to LLVM's auto-vectorization. Concluding,
we summarize VPlan, its current state and potential future.

@@

\section{Background}
\label{sec:background}

\subsection{Vectorization Strategies}
\label{back:vec-strat}
In general, vectorization techniques can be separated into different categories:

\paragraph{Inner Loop Vectorization}
Innermost loops are loops whose body's control flow does not contain any more loops.
Vectorization of these loops is more straightforward as the only option to vectorize is over 
the loop's induction variable.

@@

\paragraph{Super-Word Parallelism (SLP) Vectorization}
SLP Vectorization vectorizes similar independent sequential instructions. This can often be seen 
after a loop has been partially or fully unrolled. The resulting similar scalar 
instructions might then be vectorized later by an SLP vectorizer.

\paragraph{Outer Loop Vectorizaton}
Outer loop vectorization focuses on control flow with nested loops. Such control flow often presents
more challenges for auto-vectorization, as one could vectorize over the outer loop,
inner loop, or a mix of both, representing a hybrid/multi-dimensional vectorization approach.
Finding the best legal vectorization can therefor be more difficult.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.39\textwidth]{images/outer-loop-vec.png}
  \caption{A code snippet of a nested loop where it would be beneficial to vectorize over the 
  outer loop. The inner loop has a trip count (6) that is too low to vectorize. Additionally,
  the memory access pattern of \texttt{y} would be consecutive compared to scattered when
  vectorizing over \texttt{i} instead of \texttt{j}.}
  \label{fig:outer-loop-vec}
\end{figure}

\paragraph{Function Vectorization}
Function vectorization refers to the vectorization of entire functions. A loop might contain a
non-inlined function call, passing data related to the iteration variable. During auto-vectorization, 
this call could be replaced with a call to a newly constructed function with the same semantic 
control flow but operating on multiple values at once through vector registers.

\subsection{Vectorization Legality}
\label{sec:vec-legal}
Not all loops can be vectorized. There are various factors that can ultimately hinder any
vectorization attempts.

@@

Not all data dependencies directly hinder any vectorization~\cite{datadepvec}. An exception can, for example, 
sometimes be made for reduction idioms, as seen in the inner loop in 
Figure~\ref{fig:outer-loop-vec}, where \texttt{sum} would be a reduction variable.

Uncountable loops can typically also not be vectorized, due to their unknown number of iterations.

Finally some instructions like integer division do inherently not have a vectorized instruction equivalent.

@@

Additionally, the vectorized code size will always be larger than the scalar equivalent. This is
due to many factors: (1) The transformed control flow often includes more instructions.
(2) As the loop trip count might not be a perfect multiple of the Vectorization Factor, a scalar
epilogue/remainder of the loop might be used to handle the tail iterations. (3) Various runtime
checks might be inserted throughout the vectorized control flow to handle, for example, pointers 
to overlapping memory regions, in which case the scalar version of the loop is executed.

\subsection{Auto-Vectorization in LLVM}
LLVM implements two different vectorizers for auto-vectorization, LoopVectorize (LV)
and SLPVectorize~\cite{llvmvec,llvmhistorystate}. Auto-Vectorization is part of LLVM's 
middle-end and runs after module simplification has taken place.
In general, these vectorizers take in scalar LLVM IR and transform it into optimized vectorized 
LLVM IR.

SLPVectorize is a SLP vectorizer working on straight-line code (non-loops). It works by analyzing
the code bottom-up, keeping track of dependencies and grouping similar independent scalar instructions
together if possible. Based on a cost model SLPVectorize then decides to vectorize these groups.

LoopVectorize was designed to focus only on vectorization of innermost loops~\cite{llvmintrvplan}. 
When starting with an initial scalable loop in IR, it would first run a legality phase to check 
whether and how it is legal to vectorize the loop. This phase would produce different artifacts, 
such as if 
and what runtime aliasing checks would be required. Then, the cost model ran next, creating more 
cost-based artifacts based on the underlying scalar IR and the legality artifacts. 
Lastly, the transformation phase would vectorize 
the IR in one go based on all the previously produced artifacts.

As a potential solution, Intel proposed a long-term effort to refactor LLVM's whole auto-vectorization
using a more modular and flexible system called Vectorization Plan (VPlan)~\cite{llvmextloopvec}.

@@

The goal of Intels initial VPlan proposal was to unite any 
auto-vectorization systems (SLP, Inner-/Outer-Loop, Whole-Function) into a single flexible
system~\cite{llvmextloopvec,llvmintrvplan,llvmvplanstate}. 
This system would use VPlans to represent different 
vectorization approaches, abstracting away from the underlying scalar loop IR by using its own 
hierarchical control-flow-graph (H-CFG). This allows for iteratively exploring different 
optimization decisions by performing transformations on this CFG without modifying any LLVM IR.
Each VPlan would be able to calculate its cost and execute itself, materializing the vectorization 
approach back into LLVM IR.

This planned systems is shown in Figure~\ref{fig:vplan-future}. At first, the Legality 
Phase would check if the scalar code would be legal to vectorize at all, and if so, 
initial VPlans would be constructed. In the following Planning Phase, these initial 
VPlans would be optimized using various VPlan-to-VPlan transformations~\cite{llvmouterloopstatus}. 
The VPlan's cost model would supervise those transformations. Finally, the best VPlan 
would be chosen and executed, modifying the underlying IR.

\begin{figure}
  \centering
  \includegraphics[width=0.475\textwidth]{images/vplan-future.png}
  \caption{The planned future architecture for auto-vectorization in LLVM. The 
  optimization step in phase 2 would iteratively apply VPlan-to-VPlan transformations.}
  \label{fig:vplan-future}
\end{figure}

\subsection{VPlan Structure}
Before explaining how far this system is already implemented in LLVM we will go over
the different elements of a \texttt{VPlan} in more detail.

The purpose of a single VPlan object is to model a candidate for vectorization. 
Such a single candidate can potentially represent multiple different vectorized loops in the end. 
For example, one VPlan might model a general vectorization approach for multiple different 
Vectorization Factors (VFs) and multiple different Unroll Factors (UFs) if applicable.

\paragraph{H-CFG}
This is achieved by introducing a new layer of abstraction from the underlying loop's scalar IR.
Each VPlan stores its own Hierarchical-CFG (H-CFG) modeling the vectorization candidate. This 
control-flow graph and its components are the main element of a single VPlan.
The nodes of this H-CFG are one of three types:

\begin{itemize}
  \item \texttt{VPBasicBlock} stores a list of \texttt{VPRecipes}, similar to 
  the LLVM IR's \texttt{BasicBlock} holding a list of IR Instructions.
  \item \texttt{VPRegionBlock} represents a subgraph with a single VP entry node and a 
  single VP exit node. A \texttt{VPRegionBlock} can be marked to replicate its subgraph
  \texttt{VF * UF} times when executing. This is useful for modeling scalarized or predicated
  instructions within the final vectorized loop body.
  \item \texttt{VPIRBasicBlock} is a special \texttt{VPBasicBlock} that wraps an underlying
  IR \texttt{BasicBlock}. These nodes are useful for modeling the scalar parts around the 
  final vectorized loops such as the loop preheader and the epilogue loop.
\end{itemize}

Figure~\ref{fig:inner-loop-scalar-ir} shows the loop part of the scalar IR for the code example
of Figure~\ref{fig:inner-loop-vec}. Comparatively, Figure~\ref{fig:inner-loop-vplan-hcfg-body}
shows the loop part of the H-CFG of a VPlan that models the vectorization approach described
in Figure~\ref{fig:inner-loop-vec}.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.475\textwidth]{images/inner-loop-vplan-hcfg-loop-body-color.png}
  \caption{The vectorized loop section of the hierarchical CFG of a VPlan modeling the 
  conditional loop depicted in Figure~\ref{fig:inner-loop-vec}. 
  Surrounding \texttt{VPIRBasicBlock}-nodes are omitted.
  Some predefined values are
  \texttt{vp<\%0>} = VF * UF; \texttt{vp<\%1>} = vector-trip-count; \texttt{vp<\%2>} = original trip-count.}
  \label{fig:inner-loop-vplan-hcfg-body}
\end{figure}

\paragraph{Recipes}

\texttt{VPRecipes} can be seen as instructions of an IR specialized for auto-vectorization. 
A single \texttt{VPRecipe} inside a VPlan is based on zero or more LLVM IR instructions of the 
underlying scalar IR and will materialize into one or more LLVM IR instructions when executed.

There are many different types of \texttt{VPRecipes}, most notably:
\begin{itemize}
  \item \texttt{VPReplicateRecipe} will replicate one IR instruction either once (\texttt{CLONE})
  or multiple times (\texttt{REPLICATE}) into the vectorized IR when executed.
  \item \texttt{VPWidenRecipe} will \texttt{WIDEN} one IR instruction into a
  vector-equivalent when executed.
  \item \texttt{VPInstruction} has no underlying IR instruction it is based on and is used 
  to model newly needed instructions for the vectorized IR. When executed it will 
  \texttt{EMIT} one or more new IR instructions.
\end{itemize}

The \texttt{VPRecipes} in Figure~\ref{fig:inner-loop-vplan-hcfg-body} have a \texttt{PREFIX}
infront of them indicating their type. Additionally instructions/recipes of the same color 
between Figure~\ref{fig:inner-loop-scalar-ir} and Figure~\ref{fig:inner-loop-vplan-hcfg-body}
are related. The only recipes in our example VPlan CFG that do not semantically model any 
instructions found in the scalar IR are from the \texttt{middle.block} (purple). This block 
is used to decide if the scalar epilogue loop is to be executed following the vectorized loop.

\paragraph{Transformations}
Another main benefit of the modular VPlan system is that most vector-optimization cases can 
be directly implemented as standalone VPlan-to-VPlan transformations. A VPlan-to-VPlan 
transformation is a static method that either transforms a given VPlan or creates a new 
VPlan based on a given VPlan. The analysis for these transformations can also be done 
directly on the VPlan via its \texttt{VPRecipes}.

\subsection{State of VPlan (LLVM 19.1.4)}
\emergencystretch 6em
In its current state, VPlan is already deeply integrated into LoopVectorize (LV) and is
used to model, optimize and execute vectorization of innermost loops~\cite{llvmvplan,llvmintrvplan,llvmvplanupdate}. 
While outer loop vectorization via VPlan has already seen quite some development, any 
outer loop vectorization features are still experimental and need to be explicitly enabled 
by passing the \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} 
feature flag~\cite{llvmouterloop,llvmouterloopstatus}.

To better understand how VPlan is already being used in LLVM, we will now follow along the 
LoopVectorize pass and summarize its key components.

\subsubsection{Phase 1: Legality}
The main part of this phase is checking whether it is even legal to optimize a given loop. 
In addition to the general vectorization legality concepts mentioned in 
Section~\ref{sec:vec-legal}, LV has a number of extra conditions to check.

Some of the most important legality checks done are:
\begin{itemize}
  \item The scalar loop's control flow must not contain indirect branches or multiple backedges.
  An indirect branch is a branch that jumps to a value evaluated at runtime. A backedge in a loop 
  is an edge from the loop back to it. 
  \item Loops with non-linear control flow must be if-convertible. This means any branch or switch 
  statement must not exit the loop, all instructions of such a block can be executed speculatively 
  (e.g. via masking)
  \item Loop-carried dependencies must be vectorizable (see Section~\ref{sec:vec-legal}). In LV 
  this means explicitly that such a value must \textit{at least} either be an induction variable, a reduction 
  variable, or a fixed-order recurrence.
  \item The loop must not contain function calls outside of standard math library calls.
  \item Types must be vectorizable. Vectorizable types are integers, floating point types, and 
  pointers.
  \item The number of runtime checks to guarantee no pointers used by recurrences overlap must not be too high.
\end{itemize}

One can find a lot more details about these legality checks and their implementations by looking 
at the LLVM codebase under \texttt{LoopVectorizationLegality::\allowbreak canVectorize}.

\subsubsection{Phase 2: Planning}
After the Legality Phase has analyzed the scalar loop, initial VPlans are constructed.
A VPlan is initialized by first creating the H-CFG skeleton building blocks like preheaders, a 
region block for the vector loop, a middle block with the potential runtime check for the scalar 
epilogue loop, and a block for the scalar loop. After the H-CFG has been created, the scalar LLVM IR 
is iterated in topological order and each scalar IR instruction is transformed into a \texttt{VPRecipe}.
After finishing modeling each scalar IR instruction into the VPlan, a number of VPlan transformations 
are executed, optimizing the initialized VPlan.

After the initial VPlans have been optimized the best Vectorization Factor is computed. This is done 
by comparing all generated VPlans for all their possible VFs and choosing the most profitable one. If 
this step computes a optimal VF of $1$, it means that any form of vectorization is not profitable.

\paragraph{Cost Model}
Currently LV and VPlan still mostly rely on the old cost model. This legacy cost model derives costs 
based on the scalar LLVM IR instructions or, if applicable, their vectorized equivalent.
This is suboptimal for VPlan as VPlans work on \texttt{VPRecipes} abstracted away from the underlying 
IR. These costs per instructions are then generally just lookup tables influenced by the underlying 
compile target.

\subsubsection{Phase 3: Execution}
After the best VPlan has been chosen it is executed (\texttt{LoopVectorizationPlanner::executePlan}).
For this VPlan some final transformations are made that depend on a specifically chosen VF and UF.
First a vectorized code skeleton is created in the IR, then the VPlan is executed. Upon execution 
the H-CFG of the VPlan is walked and each node is executed, which in turn executes all of its 
\texttt{VPRecipes} in order.

Figure~\ref{fig:inner-loop-vec-cfg} shows the general CFG of a vectorized loop in LLVM and for branches 
the general branch condition per block.

\begin{figure}
  \centering
  \includegraphics[width=0.35\textwidth]{images/inner-loop-vec-loop-cfg-color.png}
  \caption{The generalized CFG for a vectorized loop with trip count \texttt{N}, 
  iteration variable \texttt{i}, vectorization factor \texttt{VF}, 
  and unroll factor \texttt{UF} in LLVM. A VPlan would explicitly model the red-highlighted 
  region in its H-CFG.}
  \label{fig:inner-loop-vec-cfg}
\end{figure}

In case that the epilogue loop could also be further vectorized, it will be in a second 
pass by creating VPlans for it, finding the best one and executing it.

@@

\subsubsection{VPlan and Integration with LV}
Setting up and integrating the VPlan structure into LV has seen tremendous effort over the last 
few years. Florian Hahn presented future directions for 
VPlan~\cite{llvmvplanupdate}, which mostly revolve around transforming more LV-based decisions into
explicit VPlan-to-VPlan transformations. Additionally, the legacy LV cost model still influences 
many cost model-related decisions. However, the transition to a VPlan-based cost model has already 
started. This new model would derive costs based on \texttt{VPRecipes} directly instead of the 
abstracted scalar IR instructions.

@@

One such project that tries to enable auto-vectorization support for inner/outer loops and 
whole-functions is the region vectorizer (RV) from the University Saarland~\cite{rv}.
The idea is that all the different vectorization strategies shown in Section~\ref{back:vec-strat}
can be generalized into a single vectorization approach.

For example, function vectorization can be transformed into loop vectorization by creating a loop 
around the function's body and vectorizing it and vice versa. 
The same applies for SLP Vectorization and loop vectorization by unrolling a loop.

RV's goal is to have such a generalized auto-vectorization by vectorizing over \textit{regions}.
A region is defined as any control flow that has a single entry and one or more non-divergent exits.
Non-divergent exits mean that starting from the same entry, control flow must not diverge to different
exits.
Such a region can then be for example an inner/outer loop, function, or SLP code~\cite{rvintro,rvproposal}.

To achieve this, RV implements partial control-flow linearization~\cite{10.1145/3192366.3192413}
and a divergence analysis that can handle unstructured CFGs. There has been an effort to partially
implement this control-flow linearization and divergence analysis into LLVM's 
upstream~\cite{rvproposal,rvproposaldep}.

@@

\section{Summary and Future Work}
\label{sec:summary}
In conclusion, VPlan has been in active development for several years now. When compiling with LLVM, 
VPlans are used to model different vectorization approaches. They do so by adding a new layer of 
abstraction over the underlying LLVM IR. This VPlan layer uses \texttt{VPRecipes} to model vectorization 
specific instructions. This abstraction allows for efficiently exploring many different optimization 
decisions at once, keeping the underlying implementation flexible and modular. In the end a cost model, 
currently still working based on LLVM IR instead of \texttt{VPRecipes} is used to compare different VPlans 
and choose the most optimal one. Finally this chosen VPlan is executed, materializing its \texttt{VPRecipes} 
back into LLVM IR.

Outer loop vectorization support via VPlan is still experimental and separated from the inner loop
LV integration. In the near future, development on outer loop vectorization will continue, and merging 
it with LV's VPlan path is planned.

@@

@@
